---
title: "Chapter 4. Exploratory Data Analysis"
subtitle: "Univariate and Multivariate techniques, and the basic principles of constructing data graphics"
format: html
---

::: {.callout-note}
## Learning Objectives (ECO Alignment)
By the end of this chapter, you will be able to:
1. Apply the "Grammar of Graphics" to create professional visualizations.
2. Perform **Univariate Analysis** to understand the distribution of a single variable.
3. Perform **Multivariate Analysis** to identify relationships between two or more variables.
4. Interpret patterns, outliers, and clusters in real-world data like `lbw.csv`.
:::

# 4.1 The Philosophy of EDA
Exploratory Data Analysis (EDA) is an iterative cycle. We use visualization and transformation to explore our data, looking for patterns or anomalies before we move on to formal statistical modeling.



# 4.2 Principles of Data Graphics: `ggplot2`
In R, we use the `ggplot2` package, which is based on the **Grammar of Graphics**. Every plot is built in layers:
1. **Data:** The dataframe (e.g., `lbw`).
2. **Aesthetics (`aes`):** The mapping of variables to visual properties (x-axis, y-axis, color).
3. **Geoms:** The geometric shapes (bars, points, lines).

# 4.3 Univariate Analysis
We begin by looking at one variable at a time. This helps us understand the "shape" of our data (skewness, central tendency, and spread).



### Numerical Variables: Histograms
```{r}
library(tidyverse)
lbw <- read_csv("data/lbw.csv")

```


# Visualizing the distribution of mother's weight (lwt)
```{r}
ggplot(lbw, aes(x = lwt)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Maternal Weight", x = "Weight (lbs)", y = "Count")
  
```


## 4.5 Tech Spotlight: Interactive Plotting with `esquisse`

While the Grammar of Graphics is powerful, it can be intimidating to remember every function name. In MAT 186, we use a tool called **esquisse**—an RStudio "Add-in" that provides a drag-and-drop interface for `ggplot2`.



### Why we use `esquisse`:
1.  **Exploration:** Quickly flip between a bar chart and a boxplot to see which represents the `lbw.csv` data better.
2.  **Code Learning:** Once you create a plot you like, `esquisse` shows you the exact `ggplot2` code used to create it. 
3.  **Professional Polish:** You can easily change colors, labels, and themes without searching through documentation.

### How to use it:
To start the tool, you can use the "Addins" menu in RStudio or run the following code in your console:
```{r}
# Install if you haven't already
# install.packages("esquisse")

# Launch the tool using our dataset
esquisse::esquisser(lbw_data)
```

