---
title: "Chapter 2. Introduction to R programming"
subtitle: "Programming in R, reading data into R, accessing R packages, writing R functions, and organizing and commenting on R code"
format: html
---

::: {.callout-note}
## Learning Objectives (ECO Alignment)
By the end of this chapter, you will be able to:
1.  **Program in R:** Understand object-oriented logic and the assignment operator.
2.  **Read Data:** Import external datasets into the R environment.
3.  **Access Packages:** Install and load specialized toolkits (The Tidyverse).
4.  **Write Functions:** Automate tasks by creating custom R functions.
5.  **Organize & Comment:** Apply professional standards for readable, reproducible code.
:::

# 2.1 Programming in R

R is an **object-oriented** language. This means we store data (numbers, text, or entire tables) in "objects" using the assignment operator `<-`. 



```{r}
# Storing a numeric vector
patient_ids <- c(101, 102, 103, 104)

# Storing a character vector
treatment_group <- c("Control", "Placebo", "Active", "Active")
```

## 2.2 Accessing R Packages

Packages are collections of functions, documentation, and sample data sets written by the R community to solve specific problems. In MAT 186, we use the **tidyverse** package as our primary toolkit. 



To use a package, you must follow a two-step process:

1.  **Installation:** This downloads the package from CRAN (the R "App Store") to your hard drive. You typically only need to do this once per computer.
2.  **Accessing (Loading):** This "opens the app" and loads the functions into your current R session. You must do this every time you restart RStudio.

### The Analogy: The Smartphone
| Step | R Command | Smartphone Equivalent |
|:---|:---|:---|
| **Step 1** | `install.packages("tidyverse")` | Downloading an App from the App Store. |
| **Step 2** | `library(tidyverse)` | Tapping the App icon to open it. |

::: {.callout-important}
## Syntax Reminder
Note that when you **install**, the package name must be in quotes: `"tidyverse"`. When you load the **library**, quotes are not required.
:::

```{r}
# Do this once
# install.packages("tidyverse")

# Do this at the start of every script
library(tidyverse)
```

## 2.3 Reading Data into R

In **Lab 5**, we move beyond simple vectors and import full datasets. Data scientists rarely type in their own data; they read it from external files. The most common format is the **CSV** (Comma Separated Values) file.

### Loading the Low Birth Weight (LBW) Dataset
For our analysis, we will use the `lbw.csv` dataset, which contains information on risk factors associated with low birth weight.



To read this file into R, we use the `read_csv()` function from the `tidyverse` library:

```{r}
library(tidyverse)

# Reading the lbw.csv data into an object named 'lbw_data'
# Assumes the file is located in a folder named 'data'
lbw_data <- read_csv("data/lbw.csv")

# Inspecting the first 6 rows of the dataset
head(lbw_data)

```

::: {.callout-note}

Understanding the Data Structure
Once imported, R treats lbw_data as a Data Frame.

Columns: Represent variables (e.g., age, lwt, smoke).

Rows: Represent individual observations (e.g., each mother in the study).
:::

**Troubleshooting the File Path**
If R gives you an error saying the file does not exist:

Ensure lbw.csv is saved in your project folder.

Check your spelling (R is case-sensitive!).

Use getwd() to verify R is looking in the correct folder.

### usin import Dataset

herramientas que lo ponen aún más fácil desde el primer día.

𝟭. 𝗨𝘀𝗮 𝗜𝗺𝗽𝗼𝗿𝘁 𝗗𝗮𝘁𝗮𝘀𝗲𝘁 𝗽𝗮𝗿𝗮 𝗮𝗽𝗿𝗲𝗻𝗱𝗲𝗿 𝗮 𝗶𝗺𝗽𝗼𝗿𝘁𝗮𝗿
La interfaz de RStudio te permite importar .csv, Excel o SPSS viendo en tiempo real cómo se genera el código.
Consejo: copia ese código a tu script para garantizar la reproducibilidad.

𝟮. 𝗨𝘀𝗮 𝗩𝗶𝗲𝘄() 𝗽𝗮𝗿𝗮 ver 𝘁𝘂𝘀 𝗱𝗮𝘁𝗼𝘀
Ordena, filtra y busca valores rápidamente.
No sustituye al código, pero permite una inspección inicial para evitar errores.

𝟯. 𝗘𝘅𝗽𝗹𝗼𝗿𝗮 𝗿á𝗽𝗶𝗱𝗼 𝗰𝗼𝗻 ggpairs()
Una sola función que genera matrices de gráficos y correlaciones automáticamente.
Analiza la estructura, relaciones y posibles problemas antes de modelar.

𝟰. 𝗔𝗽𝗿𝗲𝗻𝗱𝗲 ggplot2 𝗰𝗼𝗻 𝗮𝘆𝘂𝗱𝗮 𝘃𝗶𝘀𝘂𝗮𝗹 de esquisse
El addin esquisse permite crear gráficos ggplot2 desde una interfaz (mediante clic) y convierte los cambios en código reproducible. Modifica colores, leyendas y títulos de forma sencilla.

## 2.4 Writing R Functions

If you need to perform the same task multiple times, you should write a function. Functions encapsulate a sequence of operations into a single command, making your code more readable and less prone to error.

A function consists of:
1.  **A Name:** What you call the function (e.g., `grams_to_lbs`).
2.  **Arguments:** The input data the function needs to work.
3.  **The Body:** The code inside the curly braces `{}` that performs the math.

### Example: Converting Birth Weight
The `lbw.csv` dataset records birth weight in grams (`bwt`). We can write a function to convert those values to pounds.



```{r}
# Defining the function
grams_to_lbs <- function(grams) {
  lbs <- grams * 0.00220462
  return(lbs)
}

# Testing the function with a single value
grams_to_lbs(3000)

# Applying the function to the 'bwt' column in our dataset
lbw_data$bwt_lbs <- grams_to_lbs(lbw_data$bwt)

```

## 2.5 Organizing and Commenting on R Code

According to our ECO, code must be well-documented and organized to ensure **Reproducibility**. A reproducible script is one that a peer (or your future self) can run and understand without needing to ask you questions.

### Best Practices for MAT 186

#### 1. The "Why" Rule of Commenting
Use the `#` symbol to write comments. Comments should explain the **logic** or **justification** for a step, not just describe the code.

* **Weak:** `library(tidyverse) # load tidyverse` (The code already says this).
* **Strong:** `library(tidyverse) # loading toolkit for data import and wrangling`

#### 2. Code Sections
Use RStudio’s section headers (`Ctrl + Shift + R`) to create a table of contents for your script. This makes long analyses of datasets like `lbw.csv` much easier to navigate.



#### 3. Consistent Naming
Always use `snake_case` (all lowercase with underscores) for your objects. 
* **Good:** `lbw_clean`, `low_birth_weight_analysis`
* **Avoid:** `LBWData`, `Data1`, `test`

### Example of an Organized Script
Here is how your script for **Lab 5** should look:

```{r}

# --- 1. Setup & Data Import ---
# Purpose: Analyze risk factors for low birth weight
library(tidyverse)
lbw <- read_csv("data/lbw.csv")

# --- 2. Data Cleaning ---
# Convert birth weight from grams to pounds for US reporting standards
grams_to_lbs <- function(g) { g * 0.0022 }
lbw$bwt_lbs <- grams_to_lbs(lbw$bwt)

# --- 3. Initial Inspection ---
# Verify that columns were imported with correct data types
glimpse(lbw)
```

::: {.callout-important}

Reproducibility Check
Before submitting any lab, always restart your R session (Ctrl + Shift + F10) and run your entire script from top to bottom to ensure it works without errors.
:::
